---
title: "Reproducible Research Course Project 2"
author: "Jose Nicolas Molina"
date: '2022-07-03'
output:
  html_document:
    df_print: paged
---

# 1. Introduction
Storms and other severe weather events can cause both public health and economic problems
for communities and municipalities. Many severe events can result in fatalities, injuries, 
and property damage, and preventing such outcomes to the extent possible is a key concern.
This project involves exploring the U.S. National Oceanic and Atmospheric Administration's 
(NOAA) storm database. This database tracks characteristics of major storms and weather 
events in the United States, including when and where they occur, as well as estimates of 
any fatalities, injuries, and property damage.

# 2. Synopsis

The purpose of this analysis is to explore the storm database of the United States
National Oceanic and Atmospheric Administration (NOAA) from 1950 to 2011. This database 
contains information on climatic events that occur in the United States, as well as 
the impacts it generates on the population.
The analysis shows that tornadoes are the meteorological events with the greatest 
impact on the health of the population, followed by excessive heat. Regarding the 
economic impacts, flash floods and thunderstorm winds are the ones with the greatest 
impact, causing losses in millions of dollars in the period under study. Likewise, 
drought causes the greatest damage to crops, seconded by floods and hail.

# 3. Data Processing

Install packages & Load libraries

Install packages

load libraries 

```{r}
library(rmarkdown)
library(knitr)
library(plyr)
```

Download the storm data to the working directory folder

 Performing the download

```{r}
temp <- tempfile()
if(!file.exists("/stormData.csv.bz2")){
        download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile="./stormData.csv.bz2")
}
```

Uncompressing and reading file

```{r}
storm <- read.csv(bzfile("stormData.csv.bz2"))
```

Data exploration

The data is pre-processed. Event types are given a specific format, as follows.

Total unique events

```{r}
length(unique(storm$EVTYPE))
str(storm)
```

Translate all letters to lowercase

```{r}
eventTypes <- tolower(storm$EVTYPE)
```

Replace all dot characters with a space

```{r}
eventTypes <- gsub("[[:blank:][:punct:]+]", " ", eventTypes)
length(unique(eventTypes))
```

Update the data frame

```{r}
storm$EVTYPE <- eventTypes
```

After the cleanup, the events were significantly reduced. For the analysis, clean event data will be used.

# 4. RESULTS

QUESTION 1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

Selection of variables of interest

```{r}
SelectVar<-c("EVTYPE","FATALITIES","INJURIES","PROPDMG", "PROPDMGEXP","CROPDMG","CROPDMGEXP")
variableStorm<-storm[SelectVar]
```

Dangerous Events with respect to Population Health

To find the event types that are most harmful to population health, the number of casualties are aggregated by the event type.

```{r}
casualties <- ddply(storm, .(EVTYPE), summarize,
                    fatalities = sum(FATALITIES),
                    injuries = sum(INJURIES))
```

Events that cause most deaths and injuries

```{r}
topTenFatalities <- head(casualties[order(casualties$fatalities, decreasing = T), ], 10)
topTenInjuries <- head(casualties[order(casualties$injuries, decreasing = T), ], 10)
```

Top ten events that caused the most deaths

Group the top ten fatalities by event type and sort in descending order

```{r}
topTenFatalities[, c("EVTYPE", "fatalities")]
```

Top ten events that caused most number of injuries

Group the top ten injuries by event type and sort in descending order

```{r}
topTenInjuries[, c("EVTYPE", "injuries")]
```

Graph of the main deaths and injuries by types of meteorological events

Plot graphs showing the top ten fatalities and injuries

```{r}
par(mfrow=c(1,2),mar=c(11,4,4,2))
barplot(topTenFatalities$fatalities,names.arg=topTenFatalities$EVTYPE,las=2,col="red",ylab="Fatalities",main="Top Ten fatalities")
barplot(topTenInjuries$injuries,names.arg=topTenInjuries$EVTYPE,las=2,col="red",ylab="Injuries",main="Top Ten Injuries")
```
Analysis: The meteorological event with the greatest impact on the population turned out to be the Tornado (with the highest number of deaths and injuries)

QUESTION 2. Across the United States, which types of events have the greatest economic consequences?

Meteorological events with the greatest economic impact

Data Exploration

When reviewing the data we find columns for property damage (PROPDMG) and crop damage (CROPDMG), in addition there are other related columns of exponents (PROPDMGEXP and CROPDMGEXP).
We will transform the columns of exponents into numerical data to facilitate the respective calculations of damage to property and crops

Property damage analysis

Property damage exponents by level will be listed and assigned with their exponent data values.
Invalid data is excluded by assigning the value '0'.
So, the property damage value is calculated by multiplying the property damage and the property exponent value.

```{r}
unique(variableStorm$PROPDMGEXP)
```

Calculation of values of the exponent property in variableStorm
```{r}
variableStorm$PROPEXP[variableStorm$PROPDMGEXP == "K"] <- 1000
variableStorm$PROPEXP[variableStorm$PROPDMGEXP == "M"] <- 1e+06
variableStorm$PROPEXP[variableStorm$PROPDMGEXP == ""] <- 1
variableStorm$PROPEXP[variableStorm$PROPDMGEXP == "B"] <- 1e+09
variableStorm$PROPEXP[variableStorm$PROPDMGEXP == "m"] <- 1e+06
variableStorm$PROPEXP[variableStorm$PROPDMGEXP == "0"] <- 1
variableStorm$PROPEXP[variableStorm$PROPDMGEXP == "5"] <- 1e+05
variableStorm$PROPEXP[variableStorm$PROPDMGEXP == "6"] <- 1e+06
variableStorm$PROPEXP[variableStorm$PROPDMGEXP == "4"] <- 10000
variableStorm$PROPEXP[variableStorm$PROPDMGEXP == "2"] <- 100
variableStorm$PROPEXP[variableStorm$PROPDMGEXP == "3"] <- 1000
variableStorm$PROPEXP[variableStorm$PROPDMGEXP == "h"] <- 100
variableStorm$PROPEXP[variableStorm$PROPDMGEXP == "7"] <- 1e+07
variableStorm$PROPEXP[variableStorm$PROPDMGEXP == "H"] <- 100
variableStorm$PROPEXP[variableStorm$PROPDMGEXP == "1"] <- 10
variableStorm$PROPEXP[variableStorm$PROPDMGEXP == "8"] <- 1e+08
```

Set '0' to invalid exponent variableStorm

```{r}
variableStorm$PROPEXP[variableStorm$PROPDMGEXP == "+"] <- 0
variableStorm$PROPEXP[variableStorm$PROPDMGEXP == "-"] <- 0
variableStorm$PROPEXP[variableStorm$PROPDMGEXP == "?"] <- 0
```

Calculate the value of material damage

```{r}
variableStorm$PROPDMGVAL <- variableStorm$PROPDMG * variableStorm$PROPEXP
```

Property Crop Damage

Crop damage exponents for each level listed out & assigned those values for the crop exponent data. 
Invalid data was excluded by assigning the value as '0'. 
Then, the crop damage value was calculated by multiplying the crop damage and crop exponent value.

```{r}
unique(variableStorm$CROPDMGEXP)
```

Calculation values for the crop exponent variableStorm

```{r}
variableStorm$CROPEXP[variableStorm$CROPDMGEXP == "M"] <- 1e+06
variableStorm$CROPEXP[variableStorm$CROPDMGEXP == "K"] <- 1000
variableStorm$CROPEXP[variableStorm$CROPDMGEXP == "m"] <- 1e+06
variableStorm$CROPEXP[variableStorm$CROPDMGEXP == "B"] <- 1e+09
variableStorm$CROPEXP[variableStorm$CROPDMGEXP == "0"] <- 1
variableStorm$CROPEXP[variableStorm$CROPDMGEXP == "k"] <- 1000
variableStorm$CROPEXP[variableStorm$CROPDMGEXP == "2"] <- 100
variableStorm$CROPEXP[variableStorm$CROPDMGEXP == ""] <- 1
```

Set '0' to invalid exponent variableStorm

```{r}
variableStorm$CROPEXP[variableStorm$CROPDMGEXP == "?"] <- 0
```

Calculating the crop damage 

```{r}
variableStorm$CROPDMGVAL <- variableStorm$CROPDMG * variableStorm$CROPEXP
```

Property Damage Summary

Set property damage by event type and sort in descending order

```{r}
propDamage <- aggregate(PROPDMGVAL~EVTYPE,data=variableStorm,FUN=sum,na.rm=TRUE)
propDamage <-propDamage[with(propDamage,order(-PROPDMGVAL)),]
propDamage <- head(propDamage,10)
print(propDamage)
```

Crop Damage Summary

Crop damage by event type, sort descending

```{r}
cropDamage <- aggregate(CROPDMGVAL~EVTYPE,data=variableStorm,FUN=sum,na.rm=TRUE)
cropDamage <- cropDamage[with(cropDamage,order(-CROPDMGVAL)),]
cropDamage <- head(cropDamage,10)
print(cropDamage)
```

Graph of the economic consequences according to the type of climatic event

Graph top ten property and crop damage

```{r}
par(mfrow=c(1,2), mar = c(11,4,4,2))
barplot(propDamage$PROPDMGVAL/(10^9),names.arg=propDamage$EVTYPE,las=2,col="red",ylab="Propperty damage(billions)",main="Top Ten Property Damages")
barplot(cropDamage$CROPDMGVAL/(10^9),names.arg=cropDamage$EVTYPE,las=2,col="red",ylab="Crop damage(billions)",main="Top Ten Crop Damages")
```

# 5. Conclusions

1. Tornadoes are responsible for the greatest number of deaths and injuries, followed by excessive heat and storm wind for injuries.

2. Floods are responsible for the most damage to property, while droughts cause the most damage to crops.
